\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-13 paper.tex
Cite clonal origin paper
Closes #318

Reconstructing a most parsimonious ARG for a given data set is NP-hard
\citep{wang2001perfect}, so parsimony-based methods resort to heuristics and
are limited to analysing at most hundreds of sequences. Hence, a number of
methods aim to balance computational efficiency with reconstruction of
``reasonable", rather than parsimonious ARGs \citep{minichiello2006mapping,
parida2008estimating, kelleher2019inferring,  speidel2019method,
zhang2023biobank}. The latter three methods, as well as the parsimony-based
method of \citet{rasmussen2022espalier}, can be used on megabase scale data and
hundreds of thousands of samples under human-like parameters.

TODO mention \citep{schaefer2021ancestral} somewhere.

@@

An alternative approach is to treat the ARG as a latent parameter to be
averaged out by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, kuhner2006lamarc, nielsen2000estimation, wang2008bayesian,
wang2009population, fallon2013acg, mahmoudi2022bayesian}.
These methods operate on representations
of the ``little ARG'', and are extremely computationally expensive, being
applicable to at most hundreds of samples consisting of tens or hundreds of
kilobases with human-like parameters. State-of-the-art methods rely on cheaper,
approximate models \citep{didelot2010inference, heine2018bridging,
hubisz2020mapping,hubisz2020inference, medina2020speeding}. The most scalable
method, \texttt{ARGWeaver} \citep{rasmussen2014genome}, can be applied to
dozens of mammal-like genomes \citep{hubisz2020inference}.

Methods to sample ARGs generate a ``cloud'' of estimates, and
\citep{kuhner2017consensus} provide an approach to generate a
set of consensus breakpoints and local trees from
such a cloud.
The approach is based on examining the recombination breakpoints
in all of the input ARGS, and including those that are
in at least $k$ of the input ARGs (with some additional
filtering criteria) in the output.
Within the resulting intervals, a consensus
local tree is then generated using standard phylogenetic methods.

>>>>

Reconstructing a most parsimonious ARG for a given data set is NP-hard
\citep{wang2001perfect}, so parsimony-based methods resort to heuristics and
are limited to analysing at most hundreds of sequences. Hence, a number of
methods aim to balance computational efficiency with reconstruction of
``reasonable'', rather than parsimonious ARGs or sampling from a
distribution under a population genetics model
\citep{minichiello2006mapping,
parida2008estimating, kelleher2019inferring,  speidel2019method,
schaefer2021ancestral, zhang2023biobank}.

@@

An alternative approach is to treat the ARG as a latent parameter to be
averaged out by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, kuhner2006lamarc, nielsen2000estimation, wang2008bayesian,
wang2009population, fallon2013acg, vaughan2017inferring, mahmoudi2022bayesian}.
These methods operate on representations
of the ``little ARG'', and are computationally expensive, being
applicable to at most hundreds of samples consisting of tens or hundreds of
kilobases with human-like parameters.
State-of-the-art methods rely on cheaper,
approximate models \citep{didelot2010inference, heine2018bridging,
hubisz2020mapping,hubisz2020inference, medina2020speeding}. The most scalable
method, \texttt{ARGWeaver} \citep{rasmussen2014genome}, can be applied to
dozens of mammal-like genomes \citep{hubisz2020inference}.

Methods to sample ARGs generate a ``cloud'' of estimates, and
\cite{kuhner2017consensus} provide an approach to generate a
set of consensus breakpoints and local trees from
such a cloud.
The approach is based on examining the recombination breakpoints
in all of the input ARGS, and including those that are
in at least $k$ of the input ARGs (with some additional
filtering criteria) in the output.
Within the resulting intervals, a consensus
local tree is then generated using standard phylogenetic methods.
